Double Standards in Judging Collective Action

### 2024-04-16 Documents/manuscript.qmd
resubmit

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In five studies (total *N* = 2,979), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In three preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own ideological position---but showed no consistent ingroup bias in terms of class, race, and gender. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
keywords: collective action, double standards, social identity theory, system justification theory, Black Lives Matter
statement: |
  This research shows that what people consider an acceptable *means* of protest depends on what the *end* of a protest is. That is, people tend to judge the *same* controversial protest actions to be more acceptable when the protesters' cause aligned with their own ideological positions. This finding helps explain the often divided response to recent protest movements.
word-count: "11,546 words"
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

@@

# Load packages
library(tidyverse); library(tidybayes); library(numform); library(glue)

@@

Protest movements often spark contention about what actions are acceptable means of protest. Reactions to Black American athletes kneeling during the national anthem to protest racist police violence are a case in point. Kneeling during the national anthem is not violent, disruptive, or illegal. And yet, only 29\% of White Americans (compared to 66\% of Black Americans) and 11\% of Republicans (compared to 59\% of Democrats) considered it appropriate for Black athletes to protest in this way [@yougov_nfl_2017]. Indeed, Black athletes who had participated in the protests faced a backlash in public opinion [@lacina_nfl_2020].<!-- Likewise, 42\% of Americans thought that, in the 2020 Black Lives Matter protests, most protesters were trying to incite violence or destroy property, even though very few protests involved any violence or property destruction [@acled_year_2021]. --> This and other examples suggest that *who* the protesters are and *what* they are protesting influences how acceptable observers judge their actions to be.

In this paper, we examine double standards in judging collective action---that is, whether observers judge the same protest actions as more or less acceptable depending on their own and the protesters' group memberships (*identity-based double standards*) or on the protesters' cause and how it aligns with their own ideological positions (*ideology-based double standards*). In two studies, we first develop an instrument of 25 controversial protest actions, based in item response theory, to capture double standards in judging collective action. In three preregistered experiments, we then use this instrument to test for double standards in judging protest actions for workers' rights, for or against defunding the police, and for or against restricting abortion. We find evidence for ideology-based, but not for identity-based, double standards: People judge the same protest actions to be more acceptable when the protesters' cause aligns with their own ideological positions---but do not show consistent ingroup bias (in terms of class, race, and gender) when judging collective action by ingroup and outgroup members.

@@

Recent research has shown that this distinction matters for how people respond to collective action. @shuman_disrupting_2021 showed that non-normative, non-violent collective action is more effective than either normative or violent collective action at gaining concessions from those most resistant to social change. @feinberg_activists_2020 demonstrated, however, that observers are less supportive of social movements that use extreme protest actions.^[In contrast, White Americans exposed to pictures of destructive Black Lives Matter protests were not, on average, any less supportive of systemic change than those exposed to pictures of peaceful protests [@teixeira_white_2022].] Similarly, @teixeira_is_2020 found that advantaged-group members perceive non-normative collective action by disadvantaged-group members as more damaging to their ingroup's social image and are thus less supportive of such action. @verkuyten_intolerance_2022 found that, in general, the public tends to be intolerant of non-normative collective action. This means that activists face the dilemma that non-normative collective action might be most effective at gaining concessions but that it also risks reducing popular support for a cause. Together, these studies show that the distinction between normative and non-normative collective action is psychologically and societally consequential.

These and other studies have, for the most part, relied on ad-hoc distinctions between what researchers themselves considered normative and non-normative collective action. In liberal democracies, researchers tend to consider actions such as signing petitions, voting in elections, or peaceful protest to be normative and actions such as blocking traffic, damaging property, or violent protest to be non-normative. But, as reactions to Black athletes kneeling in protest suggest, the same action might be considered acceptable by some and unacceptable by others. That is, the distinction between normative and non-normative collective action might be in the eye of the beholder.

In this research, we move this distinction from the researcher's intuition into the realm of scientific investigation and test for double standards in where people draw the line between acceptable (normative) and unacceptable (non-normative) forms of collective action. For that purpose, we define a double standard as judging the same protest actions as more or less acceptable depending on who the protesters are and what they are protesting [for an analogous definition, see @foschi_double_2000].^[@feinberg_activists_2020 provided incidental evidence for double standards in judging collective action as, in a manipulation check, Black participants rated Black Lives Matter protests as less extreme than White participants and liberal participants rated protest actions for progressive causes as less extreme than conservative participants.]

### Identity-Based Double Standards

@@

One domain in which ingroup bias manifests is judgments about other people's actions. @hewstone_ultimate_1990 reviewed research showing that people make ingroup-serving causal attributions when judging actions by ingroup and outgroup members, for example, by attributing negative behavior by outgroup members to internal causes but attributing negative behavior by ingroup members to external causes [see @pettigrew_ultimate_1979]. @valdesolo_moral_2007 demonstrated that participants judged the same selfish action as less unfair when performed by themselves or an ingroup member than when performed by an outgroup member. Other studies [@abrams_double_2013; @endevelt_everyone_2021; for exceptions, see @mendoza_for_2014; @pinto_membership_2010] provided further evidence for ingroup bias in judging moral transgressions by ingroup and outgroup members.

In the same vein, we propose that ingroup bias results in identity-based double standards in judging collective action by ingroup and outgroup members. That is, we hypothesize that people will judge the same protest actions as more acceptable if the protesters are ingroup members (*Hypothesis 1a*) or if the cause of the protest aligns with the ingroup's interests (*Hypothesis 1b*). This hypothesis could explain why, for example, far more Black than White Americans considered it appropriate for Black athletes to kneel down to protest police violence against Black Americans [@yougov_nfl_2017]. Notably, this hypothesis applies equally to historically disadvantaged groups mobilizing against social injustice and to historically advantaged groups defending their group's position.

@@

System Justification Theory [@jost_decade_2004; for a recent review, see @jost_theory_2020] argues that people are motivated to defend, justify, and bolster the prevailing social, economic, and political system because doing so serves basic epistemic, existential, and relational needs. Both advantaged and disadvantaged groups are thought to be motivated to justify the existing system, although the strength of this motivation and its expression vary across individuals and situations.

@@

Just as system justification affects support for, and opposition to, collective action, we propose that it results in ideology-based double standards in what actions people consider acceptable forms of collective action. First, we hypothesize that  people will judge the same protest actions as more acceptable if the protest supports, rather than challenges, the system (*Hypothesis 2a*). This hypothesis reflects the assumption that the motivation to justify the system is, to some extent, universal. Second, we hypothesize that people will judge the same protest actions as more acceptable if they endorse system-justifying beliefs and the protest supports the system or if they reject system-justifying beliefs and the protest challenges the system (*Hypothesis 2b*). Notably, this hypothesis implies a symmetry in how people with different ideological orientations judge system-supporting and system-challenging collective action.

As system-justifying beliefs are closely related to political conservatism [@jost_political_2003], this hypothesis could explain partisan differences in what actions liberals and conservatives consider acceptable means of protest for progressive and conservative causes---for example, why far more Democrats than Republicans considered it appropriate to kneel during the national anthem to protest racist police violence [@yougov_nfl_2017].

### Are Partisan Double Standards Based in Ideology or Identity?

Partisan differences can result from two social psychological processes. On the one hand, political partisanship reflects ideological differences in resistance to social change and acceptance of inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated cognition, ideological differences shape how we think and feel about issues, actions, and events. On the other hand, political partisanship functions as a social identity [@finkel_political_2020; @osborne_psychology_2022] with evidence suggesting that partisans' policy preferences depend more on the stated position of their party than policy content [@cohen_party_2003]. Like other identities, political identities result in group-serving biases in how we evaluate ingroup and outgroup members. In practice, the two processes are often difficult to disentangle.

For now, we use "identity-based" to refer to double standards based on assigned social categories such as class, race, and gender and "ideology-based" to refer to double standards based on ideological alignment. We compare the evidence for identity and ideology processes in the General Discussion.

@@

```{=latex}
\begin{figure}
\centering
\caption{Item response curves for three hypothetical protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-1}
\caption*{\textit{Note.} $\theta_{(z)}$ follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top).}
\label{fig:f1}
\end{figure}
```

In this research, item response theory serves two purposes. First, item response theory provides a statistical framework for scale development. In Studies 1 and 2, we select the most discriminating and informative actions to build an instrument for capturing double standards in judging collective action. Second, item response theory provides a formal and statistical definition of double standards in judging collective action as respondents being more accepting of the same protest actions depending on who the protesters are and what they are protesting. In Experiments 1--3, we operationalize double standards as the differences in $\theta_{(z)}$ between conditions which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes.

@@

The first purpose of our research is to develop an instrument to measure where people draw the line between acceptable and unacceptable forms of collective action. In Study 1, we ask participants to generate more and less extreme protest actions and compile a pool of protest actions for scale development. In Study 2, we ask participants to rate how acceptable they consider each of those protest actions to be and use item response theory to select the most discriminating and informative protest actions for our instrument. In so doing, we build an instrument that enables both present and future research to capture double standards in judging collective action. <!-- In this way, our research situates the distinction between normative and non-normative collective action in the eye of the beholders and opens it to scientific investigation. -->

The second purpose of our research is to test for double standards in judging collective action---that is, whether people judge the same protest actions as more acceptable if the protesters' are ingroup members or the protesters' cause aligns with the ingroup's interests (*Hypothesis 1a*, *1b*) or if the protesters' cause aligns with their own system-justifying or system-challenging motives (*Hypotheses 2a*, *2b*).

In Experiment 1, we use the instrument developed in Studies 1 and 2 to test for double standards in how acceptable participants consider collective action for workers' rights in the United Kingdom. By varying both the participants' and the protesters' group memberships (working/middle class), Experiment 1 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*). Extending Experiment 1, Experiment 2 considers both system-challenging and system-defending collective action in the context of the 2020 Black Lives Matter protests in the United States. By varying the protesters' cause (for/against defunding the police) as well as the participants' and the protesters' group memberships (Black/White), Experiment 2 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*, *2b*). Building on Experiments 1--2, Experiment 3 examines how liberal and conservative men and women judge collective action for or against restricting abortion in anticipation of the Supreme Court overturning its *Roe v. Wade* decision on reproductive rights in the United States. By varying the protesters' cause as well as the participants' ideology and gender, Experiment 3 compares the relative evidence for identity-based (*Hypothesis 1b*) and ideology-based (*Hypothesis 2b*) double standards in judging collective action concerning a conservative cause in a sample balanced by gender and ideology. Together, the preregistered experiments provide a comprehensive test of the hypothesized double standards in judging collective action.

@@

@@

## Study 1

In Study 1, we compiled protest actions from participants' responses and other sources. We recruited 60 participants from the Prolific subject pool, all of whom were citizens of the UK or the US. To increase the socioeconomic diversity of our sample, we recruited 30 non-students without a university degree, 15 non-students with a university degree, and 15 current university students.

Participants first read an accessible definition of what a social group is and that collective action is any action group members take to promote the interests of their social group. Participants were then asked to name between five and ten actions that fit that definition. Participants were encouraged to think of actions that varied in how acceptable they were in their opinion. We recoded responses into a smaller set of unique actions, which we supplemented with protest actions from the psychological and political science literature [e.g., @sharp_nonviolent_1973]. This process resulted in 72 actions that varied in how acceptable we would expect them to be. For details, see Supplemental Online Material (SOM).

## Study 2

@@

We estimated a graded response model [@burkner_bayesian_2021; @samejima_graded_1997]---an item response theory model for ordinal response variables---for participants' ratings of how often an action would be an acceptable form of collective action.^[Ratings of how often a given action would be an acceptable form of collective action were strongly and negatively correlated with how disruptive, violent, and extreme participants considered the action to be and how negative they felt about it (Table S3).] Based on the results, we selected the 25 most discriminating, informative, and relevant protest actions to form an instrument for measuring double standards in judging collective action in Experiments 1 and 2. We selected controversial, and thus diagnostic, actions (e.g., disrupting traffic) over actions that almost all respondents considered acceptable (e.g., signing petitions) or unacceptable (e.g., political violence; Figure \ref{fig:f2}). For detailed ratings and results, see SOM.

@@

In Experiment 1, we used the newly developed instrument to test for double standards in judging collective action for workers' rights in the United Kingdom. That is, we tested whether people with either working-class or professional jobs applied different standards when judging collective action taken by people with either working-class or middle-class jobs to protest against a fictitious government bill threatening their rights.

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that, in line with *Hypothesis 1a* and *1b*, participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to working-class and middle-class protesters, we tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (middle-class) group members. Assuming that collective action by lower-status group members is seen as more threatening to the unequal status quo than collective action by higher-status group members and that people are, to same extent, motivated to justify the system, we tested the preregistered hypothesis that, in line with *Hypothesis 2a*, participants would judge the same protest actions as more acceptable if the protesters were from a higher-status group than if they were from a lower-status group. In non-preregistered analyses, we further explored whether participants would judge the same protest actions as more or less acceptable depending on their own ideological positions. In this way, Experiment 1 provided a first test of the hypothesized identity- and ideology-based double standards in judging collective action.

@@

We used a screening survey to recruit participants who satisfied our preregistered inclusion criteria for the lower-status and higher-status groups. For the lower-status group, we recruited 475 participants who did not have a university degree and placed themselves on the bottom three ranks of a subjective socioeconomic status ladder. For the higher-status group, we recruited 400 participants who had at least an undergraduate degree and placed themselves on the top four ranks of the subjective socioeconomic status ladder.

@@

We recruited 500 participants from the remaining 687 participants, 250 from the lower-status group and 250 from the higher-status group. Participants were randomly assigned to read a vignette about a government bill affecting either people in working-class jobs (lower-status protesters) or people in professional jobs (higher-status protesters). Participants in both conditions were instructed to carefully read the vignette and to try to imagine what it would be like if this situation was real. Participants in the lower-status protesters condition read the following introduction:

>>>>

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: |
    Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. Five studies (total $N = 2,979$) investigated double standards in judging collective action---that is, whether observers judge the same protest actions to be more acceptable depending on who the protesters are and what they are protesting. Two studies used item response theory to develop an instrument of 25 controversial protest actions to measure where people draw the line between acceptable and unacceptable forms of collective action. Three preregistered experiments found no consistent evidence for ingroup bias in terms of social class when judging protests for workers' rights (Experiment 1), in terms of race when judging protests for and against defunding the police (Experiment 2), and in terms of gender when judging protests for and against restricting abortion (Experiment 3). Instead, we found that participants who identified as left-wing/liberal (Experiments 1--3) and rejected system-justifying beliefs (Experiments 1--2) considered the same protest actions more acceptable when a cause aligned with their ideological orientation (for workers' rights, for defunding the police, against restricting abortion) than when it did not (against defunding the police, for restricting abortion). Conservative participants considered the same actions somewhat more acceptable when protesters supported, rather than opposed, restricting abortion (Experiment 3) but considered all protest actions, for and against defunding the police, equally unacceptable (Experiment 2). Our findings have theoretical and practical implications for understanding the often-divided response to social movements.
keywords: collective action, double standards, social identity theory, system justification theory, partisanship, Black Lives Matter, abortion
statement: |
  This research shows that what people consider an acceptable *means* of protest depends on what the *end* of a protest is. That is, observers sometimes judge the *same* controversial protest action (e.g., blocking a road) to be more acceptable when the protesters' cause (e.g., defunding the police) aligns with their own ideological position. Conservatives, who support social order and oppose social change, tend to consider all protest actions less acceptable than progressives. Progressives, in turn, tend to consider disruptive but non-violent actions acceptable means to achieve social change and challenge social injustice. This finding helps explain the often-divided response to recent protest movements.
word-count: "11,654 words"
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  docx:
    reference-doc: word-template.docx
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

@@

  # Load packages
  library(tidyverse)
  library(tidybayes)
  library(numform)
  library(glue)

@@

Protest movements often spark contention about what actions are acceptable means of protest. Reactions to Black American athletes kneeling during the national anthem to protest racist police violence are a case in point. Kneeling during the national anthem is not violent, disruptive, or illegal. And yet, only 29\% of White Americans (compared to 66\% of Black Americans) and 11\% of Republicans (compared to 59\% of Democrats) considered it appropriate for Black athletes to protest in this way [@yougov_nfl_2017].<!-- Indeed, Black athletes who had participated in the protests faced a backlash in public opinion [@lacina_nfl_2020].--> <!--Likewise, 42\% of Americans thought that, in the 2020 Black Lives Matter protests, most protesters were trying to incite violence or destroy property, even though very few protests involved any violence or property destruction [@acled_year_2021]. -->This and other examples suggest that *who* the protesters are and *what* they are protesting influences how acceptable their actions are judged to be.

In this paper, we examine double standards in judging collective action---that is, whether observers judge the *same* protest actions as more or less acceptable depending on their own and the protesters' group memberships *or* on the protesters' cause and how it aligns with their own ideological positions.<!-- In two studies, we first develop an instrument of 25 controversial protest actions, based in item response theory, to capture double standards in judging collective action. In three preregistered experiments, we then use this instrument to test for double standards in judging protest actions for workers' rights, for or against defunding the police, and for or against restricting abortion. We do not find consistent evidence for ingroup bias in terms of class, race, and gender when judging collective action by ingroup and outgroup members. Instead, we find that liberals and, less so, conservatives judge the same protest actions to be more acceptable when the protesters' cause aligns with their own ideological positions. -->

@@

Recent research has shown that this distinction matters for how people respond to collective action [for a review, see @shuman_when_2023]. @shuman_disrupting_2021 showed that non-normative, non-violent collective action is more effective than either normative or violent collective action at gaining concessions from those most resistant to social change. @feinberg_activists_2020 demonstrated, however, that observers are less supportive of social movements that use extreme protest actions.^[In contrast, White Americans exposed to pictures of destructive Black Lives Matter protests were not, on average, any less supportive of systemic change than those exposed to pictures of peaceful protests [@teixeira_white_2022].] Similarly, @teixeira_is_2020 found that advantaged-group members perceive non-normative collective action by disadvantaged-group members as more damaging to their ingroup's social image and are thus less supportive of such action. This means that activists face the dilemma that non-normative collective action might be most effective at gaining concessions but that it also risks reducing popular support for a cause. Together, these studies show that the distinction between normative and non-normative collective action is psychologically and societally consequential.

Past research has, for the most part, relied on ad-hoc distinctions between what researchers themselves considered normative and non-normative collective action. In liberal democracies, researchers tend to consider actions such as signing petitions, voting in elections, or peaceful protest to be normative and actions such as blocking traffic, damaging property, or violent protest to be non-normative. But, as reactions to Black athletes kneeling in protest suggest, the same action might be considered acceptable by some and unacceptable by others.<!-- That is, the distinction between normative and non-normative collective action might be in the eye of the beholder. --> In this research, we <!--move this distinction from the researcher's intuition into the realm of scientific investigation and -->test for double standards in where people draw the line between what is an acceptable (normative) and an unacceptable (non-normative) form of collective action.^[In this research, we use *acceptable* ["agreed or approved of by most people in a society", @acceptable] as a plain-language synonym for *normative* ("conforming to the norms of the existing social system").] For that purpose, we define a double standard as judging the same protest actions as more or less acceptable depending on who the protesters are and what they are protesting [for a similar definition, see @foschi_double_2000].^[@feinberg_activists_2020 provided incidental evidence for double standards in judging collective action as, in a manipulation check, Black participants rated Black Lives Matter protests as less extreme than White participants and liberal participants rated protest actions for progressive causes as less extreme than conservative participants.]

### Group-Based Double Standards

Far more Black ($66\%$) than White ($29\%$) Americans considered it appropriate for Black athletes to kneel during the national anthem to protest racist police violence [@yougov_nfl_2017]. This suggests that what protest actions people consider acceptable might depend on their own and the protesters' group memberships.

@@

One domain in which ingroup bias manifests is judgments about other people's actions. @hewstone_ultimate_1990 reviewed research showing that people make ingroup-serving causal attributions when judging actions by ingroup and outgroup members. For example, people tend to attribute negative behavior by outgroup members to internal causes but attribute negative behavior by ingroup members to external causes [see @pettigrew_ultimate_1979]. @valdesolo_moral_2007 demonstrated that participants judged the same selfish action as less unfair when performed by themselves or an ingroup member than when performed by an outgroup member. Other studies [@abrams_double_2013; @endevelt_everyone_2021; for exceptions, see @mendoza_for_2014; @pinto_membership_2010] provided further evidence for ingroup bias in judging moral transgressions by ingroup and outgroup members.

In the same vein, we propose that ingroup bias results in double standards in judging collective action by ingroup and outgroup members. That is, we hypothesize that people will judge the same protest actions as more acceptable if the protesters are ingroup members and if the cause of the protest aligns with the ingroup's interests. Notably, this hypothesis applies equally to historically disadvantaged groups mobilizing against social injustice and to historically advantaged groups defending their group's position.<!-- In this research, we investigate group-based double standards in terms of class, race, and gender. -->

@@

Far more Democrats ($59\%$) than Republicans ($11\%$) considered it appropriate to kneel during the national anthem to protest racist police violence [@yougov_nfl_2017]. This suggests that what protest actions people consider acceptable might depend on the protesters' cause and how it aligns with their own ideological positions.

One explanation for ideology-based double standards is that they result from the same intergroup processes as other group-based double standards. In this view, political partisanship functions as a social identity [@finkel_political_2020; @osborne_psychology_2022] and results from universal cognitive mechanisms for forming, detecting, and maintaining alliances [@pinsof_strange_2023]. Like other identities, political identities result in group-serving biases in how we evaluate ingroup and outgroup members.<!-- Likewise, political partisanship might result from universal cognitive mechanisms for forming, detecting, and maintaining alliances [@pinsof_strange_2023] which, in turn, cause biases that serve to support political allies and oppose political rivals.--> This explanation aligns with Verkuyten et al.'s [-@verkuyten_intolerance_2022] finding that participants were more tolerant of transgressive protest actions (e.g., sending hate mail) when taken by their most-liked, rather than their least-liked, political group (e.g., climate activists).

Another explanation for ideology-based double standards is that they are rooted in ideological differences in how conservatives and progressives relate to social change and social inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated social cognition, ideological differences shape how we think and feel about political issues, actions, and events. In this view, ideology-based double standards depend on whether the cause of a protest aligns with someone's ideological positions.

System Justification Theory [@jost_decade_2004; for a recent review, see @jost_theory_2020], in particular, argues that people are motivated to defend, justify, and bolster the prevailing social, economic, and political system because doing so serves basic epistemic, existential, and relational needs. Both advantaged and disadvantaged groups are thought to be motivated to justify the existing system, although the strength of this motivation and its expression vary across individuals and situations. Conservatives, far more so than progressives, endorse system-justifying beliefs.

@@

Just as system justification affects support for, and opposition to, collective action, we propose that it results in ideology-based double standards in what actions people consider acceptable forms of collective action. That is, we hypothesize that people will judge the same protest actions as more acceptable if they endorse system-justifying beliefs and the protest supports the system or if they reject system-justifying beliefs and the protest challenges the system. <!-- Notably, this hypothesis implies a symmetry in how people with different ideological orientations judge system-supporting and system-challenging collective action.  -->

<!-- What both explanations have in common is [ingroup bias]. [This perspective would explain ideology-based double standards as a group-based double standard.]  [Ingroup bias, in this account, is symmetric.] -->

<!-- [In this view, the content of those coalitions is arbitrary and a historical accident.] [Cohen, 2003] -->

<!-- A related perspective ... What both perspectives have in common, is that the content of ... and its effects are symmetric. -->

<!-- When we understand political partisanship as an identity or alliance, we might understand ideology-based double standards as an expression of ingroup bias based on partisan identities and alliances.  -->

<!-- Partisan differences can result from two social psychological processes. On the other hand, political partisanship functions as a social identity [@finkel_political_2020; @osborne_psychology_2022] with evidence suggesting that partisans' policy preferences depend more on the stated position of their party than policy content [@cohen_party_2003]. Like other identities, political identities result in group-serving biases in how we evaluate ingroup and outgroup members. In this view, ... -->

<!-- In practice, the two processes are often difficult to disentangle. -->

<!-- One explanation is that political partisanship reflects ideological differences in resistance to social change and acceptance of inequality which, in turn, are motivated by epistemic, existential, and relational needs (Jost et al., 2003; Jost, 2021). In particular, SJT ... -->

<!-- On the one hand, political partisanship reflects ideological differences in resistance to social change and acceptance of inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated cognition, ideological differences shape how we think and feel about issues, actions, and events.  -->
<!-- This perspective, however, assumes that ideologies are essentially arbitrary in content and symmetric. OR While this perspective, on its own, might explain double standards, it is incomplete. At the same time, there are real ... OR By assuming that political ideologies are arbitrary and symmetric, this perspective ... OR Individual Differences and Needs OR not SUFFICIENT TO EXPLAIN -->

In principle, the two explanations for ideology-based double standards---in terms of intergroup processes and ideological differences---could be disentangled by separating political identities and causes. An explanation in terms of intergroup processes posits that political identities, not causes, shape partisan behavior. Indeed, research suggests that partisans' policy preferences depend more on the stated position of their party than policy content [@cohen_party_2003]. An explanation in terms of ideological differences instead posits that there is an "elective affinity" between political causes and the psychological needs and motives of their supporters [@jost_left_2021]. In other words, the content of political ideologies is not arbitrary or a historical accident, as implied by an explanation in terms of intergroup processes [see @pinsof_strange_2023]. In practice, however, the two explanations are difficult to disentangle in the context of real-world protest movements in which political identities and causes can seldom be separated.

For now, we use "group-based" to refer to double standards based on ascribed categories such as class, race, and gender and "ideology-based" to refer to double standards based on ideological alignment. We compare the evidence for different explanations of those double standards in the General Discussion.

@@

```{=latex}
\begin{figure}
\centering
\caption{Item response curves for three hypothetical protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-1.png}
\caption*{\textit{Note.} $\theta_{(z)}$ follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top).}
\label{fig:f1}
\end{figure}
```

In this research, item response theory serves two purposes. First, it provides a statistical framework for scale development. In Studies 1 and 2, we select the most discriminating and informative actions to build an instrument for capturing double standards in judging collective action. Second, it provides a formal and statistical definition of double standards in judging collective action as respondents being more accepting of the same protest actions depending on who the protesters are and what they are protesting. In Experiments 1--3, we operationalize double standards as differences in $\theta_{(z)}$ which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes.

@@

Five studies (total $N = 2,979$) investigate double standards in judging collective actions---that is, whether observers judge the same controversial protest actions to be more acceptable depending on who the protesters are and what they are protesting.

Our first purpose is to develop an instrument to measure where people draw the line between normative (acceptable) and non-normative (unacceptable) forms of collective action. In Study 1, we ask participants to generate more and less extreme protest actions and compile a pool of protest actions for scale development. In Study 2, we ask participants to rate how acceptable they consider each of those protest actions to be and use item response theory to select the most discriminating and informative protest actions for our instrument. In so doing, we build an instrument that enables both present and future research to capture double standards in judging collective action. <!-- In this way, our research situates the distinction between normative and non-normative collective action in the eye of the beholders and opens it to scientific investigation. -->

Our second purpose is to investigate potential double standards in judging collective action. Three preregistered experiments used the newly developed instrument to compare how acceptable participants controversial protest actions in different conditions.

By varying the participants' and the protesters' group memberships, we test for ingroup bias in terms of social class when judging protests for workers' rights (Experiment 1), in terms of race when judging protests for and against defunding the police (Experiment 2), and in terms of gender when judging protests for and against restricting abortion (Experiment 3). In this way, we test the hypothesis that people will judge the same protest actions as more acceptable if the protesters are ingroup members or the cause of the protest aligns with the ingroup's interests.

By varying the protesters' causes, we test whether participants who identify as left-wing/liberal (Experiments 1--3) and reject system-justifying beliefs (Experiments 1--2) judge the same protest actions to be more acceptable when the protesters' cause is progressive and system challenging (for workers' rights, for defunding the police, against restricting abortion). Likewise, we test whether participants who identify as right-wing/conservative and endorse system-justifying beliefs judge the same protest actions to be more acceptable when the protesters' cause is conservative and system defending (against defunding the police, against restricting abortion). In this way, we test the hypothesis that people will judge the same protest actions as more acceptable if the cause of the protest aligns with their own ideological positions.

<!-- In Experiment 1, we use the instrument developed in Studies 1 and 2 to test for double standards in how acceptable participants consider collective action for workers' rights in the United Kingdom. By varying both the participants' and the protesters' group memberships (working/middle class), Experiment 1 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*). Extending Experiment 1, Experiment 2 considers both system-challenging and system-defending collective action in the context of the 2020 Black Lives Matter protests in the United States. By varying the protesters' cause (for/against defunding the police) as well as the participants' and the protesters' group memberships (Black/White), Experiment 2 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*, *2b*). Building on Experiments 1--2, Experiment 3 examines how liberal and conservative men and women judge collective action for or against restricting abortion in anticipation of the Supreme Court overturning its *Roe v. Wade* decision on reproductive rights in the United States. By varying the protesters' cause as well as the participants' ideology and gender, Experiment 3 compares the relative evidence for identity-based (*Hypothesis 1b*) and ideology-based (*Hypothesis 2b*) double standards in judging collective action concerning a conservative cause in a sample balanced by gender and ideology. Together, the preregistered experiments provide a comprehensive test of the hypothesized double standards in judging collective action. -->

@@

In Study 1, we compiled protest actions from participants' responses and other sources. We recruited 60 participants from the Prolific subject pool, all of whom were citizens of the UK or the US. To increase the socioeconomic diversity of our sample, we recruited 30 non-students without a university degree, 15 non-students with a university degree, and 15 current university students. Participants first read an accessible definition of what a social group is and that collective action is any action group members take to promote the interests of their social group.

> Society is not only made up of individuals, but consists of 'social groups' to which these individuals belong. Each person lives in a place, has a job (or not) at an organisation, is a fan of a specific sports club, has a religion, or belongs to any number of other such groups. Individuals often act in ways to promote the interests of the social groups to which they belong.

\noindent Participants were then asked to name between five and ten actions that fit that definition. Participants were encouraged to think of actions that varied in how acceptable they were in their opinion. We recoded responses into a smaller set of unique actions, which we supplemented with protest actions from the psychological and political science literature [e.g., @sharp_nonviolent_1973]. This process resulted in 72 actions that varied in how acceptable we would expect them to be. For details, see Supplemental Online Material (SOM).

@@

@@

We estimated a graded response model [@burkner_bayesian_2021; @samejima_graded_1997]---an item response theory model for ordinal response variables---for participants' ratings of how often an action would be an acceptable form of collective action.^[Ratings of how often a given action would be an acceptable form of collective action were strongly and negatively correlated with how disruptive, violent, and extreme participants considered the action to be and how negative they felt about it (Table S3).] Based on the results, we selected the 25 most discriminating, informative, and relevant protest actions to form an instrument for measuring double standards in judging collective action in Experiments 1--3. We selected controversial, and thus diagnostic, actions (e.g., disrupting traffic) over actions that almost all respondents considered acceptable (e.g., signing petitions) or unacceptable (e.g., political violence; Figure \ref{fig:f2}). For detailed ratings and results, see SOM.

@@

In Experiment 1, we used the newly developed instrument to test for double standards in judging collective action for workers' rights in the United Kingdom. That is, we tested whether people with either working-class or professional jobs applied different standards when judging collective action taken by people with either working-class or middle-class jobs to protest against a fictitious government bill threatening their group's rights.

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to working-class and middle-class protesters, we further tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (middle-class) group members.^[For Experiments 1 and 2, we had preregistered the additional hypotheses that all participants would  judge  the  same actions  as  more  acceptable when the protesters were from the higher-status group rather than the lower-status group (Experiment 1) and when the goal of the protest was to support rather than challenge the system (Experiment 2). These hypotheses reflected the assumption that the motivation to justify the system was, to some extent, universal. With hindsight, we no longer think that this assumption is tenable and, therefore, omit the hypotheses based on it from the revised manuscript. We still report the same preregistered analyses and their results.] In non-preregistered analyses, we tested whether participants would judge the same protest actions as more or less acceptable depending on their own ideological positions in terms of political orientation, system-justifying beliefs, and social dominance orientation.^[In Experiment 1, we explored a broader range of operationalizations of political ideology, including social dominance orientation, than in the other experiments.] In this way, Experiment 1 provided a first test of the hypothesized class- and ideology-based double standards in judging collective action.

@@

We used a screening survey to recruit participants who satisfied our preregistered inclusion criteria for the lower-status and higher-status groups. For the lower-status group, we recruited participants who did not have a university degree and placed themselves on the bottom three ranks of a subjective socioeconomic status ladder. For the higher-status group, we recruited participants who had at least an undergraduate degree and placed themselves on the top four ranks of the subjective socioeconomic status ladder.

@@

We recruited 500 participants from the remaining participants, 250 from the lower-status group and 250 from the higher-status group. Participants were randomly assigned to read a vignette about a government bill affecting either people in working-class jobs (lower-status protesters) or people in professional jobs (higher-status protesters). Participants in both conditions were instructed to carefully read the vignette and to try to imagine what it would be like if this situation was real. Participants in the lower-status protesters condition read the following introduction:
